Reagents: expressing and composing fine-grained concurrency

### 2011-10-14 paper/paper.tex
expanding background section

\newcommand{\elide}[1]{}

@@

Broadly, we are interested in data structures and algorithms for
communication, synchronization, or both.  This section gives a brief
survey of the most important technqiues for communication and
synchronization in a fine-grained setting---exactly the techniques
that reagents abstract and generalize.  Readers already familiar with
Treiber stacks~\cite{?}, elimination-backoff stacks~\cite{?}, dual
stacks~\cite{?}, and MCS locks~\cite{?} can safely skip to
\secref{reagents}.

Given our target of cache-coherent, shared-memory architectures, the
most direct way of communicating between threads is modifying shared
memory.  The challange is to provide both \emph{atomicity} and
\emph{scalability}: communications must happen concurrently both
without corruption and without clogging the limited memory bandwidth,
even when many cores are communicating.  A simple way to provide
atomicity is to associate a lock with each shared data structure,
acquiring the lock before performing any operation.  The prevailing
wisdom\footnote{With some notable exceptions, like the recent work on
  \emph{flat combining}~\cite{?}, which we explore in \secref{?}} is
that such \emph{coarse-grained} locking is inherently unscalable:
\begin{itemize}
\item It forces operations on the data structure to be serialized,
  even when they could be performed in parallel.
\item It adds extra cache-coherence traffic, since each core must
  acquire the same lock's cache line in exclusive mode before
  operating.  For fine-grained communication, that means at least one
  cache miss per operation.
\item It is susceptible to ... % blocking 
\end{itemize}

@@

traditional code for Treiber stack (with exp. backoff),

\begin{figure}
\begin{lstlisting}
class TreiberStack[A] {
  private val head = new AtomicReference[List[A]](Nil)
  def push(a: A) {
    val backoff = new Backoff
    while (true) {
      val cur = head.get()
      if (head.cas(cur, a :: cur)) return
      backoff.once()
    }
  }
  def tryPop(): Option[A] = {
    val backoff = new Backoff
    while (true) {
      val cur = head.get() 
      cur match {
        case Nil     => return None
        case a::tail => if (head.cas(cur, tail)) return a
      }
      backoff.once()
    }
  }
}
\end{lstlisting}
\caption{Treiber's stack}
\end{figure}

need to explain pattern matching and PartialFunction

talk about elimination backoff and blocking versions but don't show code.

@@

\begin{itemize}
\item TreiberStack
\item with blocking
\item with elimination backoff
\item MSQueue
\item with blocking
\item with buckets?
\item Counter -> semaphore
\item RWLocks
\item Tree barriers
\item Dining philosophers(ish)
\item Flat combining
\end{itemize}

>>>>

\newcommand{\secref}[1]{\S\ref{sec:#1}}
\newcommand{\figref}[1]{Fig.~\ref{fig:#1}}
\newcommand{\elide}[1]{}

@@

Broadly, we are interested in data structures and algorithms for
communication, synchronization, or both.  This section gives a brief
survey of the most important techniques for communication and
synchronization in a fine-grained setting---exactly the techniques
that reagents abstract and generalize.  Readers already familiar with
Treiber stacks~\cite{?}, elimination-backoff stacks~\cite{?}, dual
stacks~\cite{?}, and MCS locks~\cite{?} can safely skip to
\secref{reagents}.

Given our target of cache-coherent, shared-memory architectures, the
most direct way of communicating between threads is modifying shared
memory.  The challenge is to provide both \emph{atomicity} and
\emph{scalability}: communications must happen concurrently both
without corruption and without clogging the limited memory bandwidth,
even when many cores are communicating.  A simple way to provide
atomicity is to associate a lock with each shared data structure,
acquiring the lock before performing any operation.  The prevailing
wisdom\footnote{With some notable exceptions, like the recent work on
  \emph{flat combining}~\cite{?}, which we explore in \secref{?}.} is
that such \emph{coarse-grained} locking is inherently unscalable:
\begin{itemize}
\item It forces operations on the data structure to be serialized,
  even when they could be performed in parallel.
\item It adds extra cache-coherence traffic, since each core must
  acquire the same lock's cache line in exclusive mode before
  operating.  For fine-grained communication, that means at least one
  cache miss per operation.
\item It is susceptible to ... % blocking %TODO
\end{itemize}

%% Say something about CONTENTION.

To combat scalability problems, data structures employ finer-grained locking,
or eschew locking altogether.  Fine-grained locking associates locks with
small, independent parts of a data structure, allowing those parts to be
manipulated in parallel.  Lockless (or \emph{nonblocking}) data structures
instead perform updates directly, using hardware-level operations (like
compare-and-set) to ensure atomicity.  Doing the updates directly means that
there is no extra communication or contention for locks, but it also generally
means that the entire update must consist of changing a single word of
memory---a constraint which is often quite challenging to meet.

\begin{figure}
\begin{lstlisting}
class TreiberStack[A] {
  private val head = new AtomicReference[List[A]](Nil)
  def push(a: A) {
    val backoff = new Backoff
    while (true) {
      val cur = head.get()
      if (head.cas(cur, a :: cur)) return
      backoff.once()
    }
  }
  def tryPop(): Option[A] = {
    val backoff = new Backoff
    while (true) {
      val cur = head.get() 
      cur match {
        case Nil     => return None
        case a::tail => 
          if (head.cas(cur, tail)) return Some(a)
      }
      backoff.once()
    }
  }
}
\end{lstlisting}
\nocaptionrule
\caption{Treiber's stack (in Scala)}
\label{fig:classic-treiber}
\end{figure}

To make this discussion concrete, we give a classic example of a fine-grained
concurrent data structure: Treiber's lock-free stack~\cite{?}, shown in
\figref{classic-treiber}.
%notice: changed to less desirable representation to make CAS work

%Most of the Scala code in this paper should be 
%need to explain pattern matching and PartialFunction

@@

Treiber's stack scales better than a lock-based stack mainly because it
decreases the amount of shared data that must be updated per operation:
instead of acquiring a shared lock, altering a shared stack pointer, and
releasing the shared lock, Treiber's stack does a single CAS to update the
shared pointer directly.  Scalability can be further improved by

%talk about elimination backoff and blocking versions but don't show code.

@@

\begin{itemize}
\item TreiberStack
\item with blocking
\item with elimination backoff
\item MSQueue
\item with blocking
\item with buckets?
\item Counter to semaphore
\item RWLocks
\item Tree barriers
\item Dining philosophers(ish)
\item Flat combining
\end{itemize}
